---
title: "Northwinds Data Analysis"
author: "Ayesha Afzaal"
date: "2024-01-21"
output: html_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#load necessary libraries
library(tidyverse)
library(zoo)
library(ggplot2)
library(readr)
library(knitr)
library(plotly)

# loading the Northwind dataset
setwd("C:/Users/dell/Desktop/IEC")
orders <- read_csv("Northwind/orders.csv")
orderdetails <- read_csv("Northwind/orderdetails.csv")
products <- read_csv("Northwind/products.csv")
suppliers <- read_csv("Northwind/suppliers.csv")
employees <- read_csv("Northwind/employees.csv")
customers <- read_csv("Northwind/customers.csv")

```

# Top Selling Product
```{r, message=FALSE, warning=FALSE}                        
# Calculate revenue for each product
product_revenue <- orderdetails |>
  inner_join(products, by = "ProductID") |>
  mutate(Revenue = orderdetails$UnitPrice * orderdetails$Quantity * 
           (1 - orderdetails$Discount))

# Identify the top N products by revenue
top_products <- product_revenue |>
  group_by(ProductName) |>
  summarise(TotalRevenue = sum(Revenue)) |>
  arrange(desc(TotalRevenue)) |>
  head(30)

# Display the Top Selling Products using DT::datatable
DT::datatable(
  top_products,
  caption = "Top Selling Products",
  options = list(
    lengthMenu = c(5, 10),
    pageLength = 5,
    searching = TRUE,
    ordering = TRUE
  ),
  style = "bootstrap",
  class = "table table-bordered table-hover"
)


```

***

# Employee-wise Sales Comparison Using Bar Plot              
```{r, message=FALSE, warning=FALSE}
# Join Employees with Orders and OrderDetails
joined_data <- inner_join(employees, orders, by = "EmployeeID") |>
  inner_join(orderdetails, by = "OrderID")

# Calculate total sales per employee
employee_sales <- joined_data |>
  group_by(EmployeeID, FirstName, LastName) |>
  summarise(TotalSales = sum(UnitPrice * Quantity * (1 - Discount)))

# Create a bar plot for employee-wise sales comparison
ggplot(employee_sales, aes(x = paste(FirstName, LastName), y = TotalSales, fill = paste(FirstName, LastName))) +
  geom_bar(stat = "identity") +
  labs(title = "Employee-wise Sales Comparison", x = "Employee", y = "Total Sales", fill = "Employee Name") +
  scale_fill_manual(values = rainbow(nrow(employee_sales)), name = "Employee Name") +  # Customize the legend label
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

***

# Supplier Contribution                            
```{r, message=FALSE, warning=FALSE} 
# Join orderdetails with products
joined_data <- orderdetails |>
  inner_join(products, by = "ProductID") |>
  inner_join(suppliers, by = "SupplierID")

# Calculate Supplier Revenue
supplier_revenue <- joined_data |>
  mutate(SupplierRevenue = orderdetails$UnitPrice * orderdetails$Quantity * 
           (1 - orderdetails$Discount))

# Group by SupplierName and calculate TotalSupplierRevenue
supplier_contribution <- supplier_revenue |>
  group_by(SupplierID, CompanyName) |>
  summarise(TotalSupplierRevenue = sum(SupplierRevenue))|>
  arrange(desc(TotalSupplierRevenue)) |>
  head(30)


# Display the Supplier Contribution using DT::datatable
DT::datatable(
  supplier_contribution,
  caption = "Supplier Contribution",
  options = list(
    lengthMenu = c(5, 10),
    pageLength = 10,
    searching = TRUE,
    ordering = TRUE
  ),
  style = "bootstrap",
  class = "table table-bordered table-hover"
)

```

***

# Order Date Analysis
```{r, message=FALSE, warning=FALSE}
# Convert OrderDate to a Date object
orders$OrderDate <- as.Date(orders$OrderDate, format = "%m/%d/%Y")

# Extract and analyze year, month, and day
orders$Year <- format(orders$OrderDate, "%Y")
orders$Month <- format(orders$OrderDate, "%m")
orders$Day <- format(orders$OrderDate, "%d")

# Analysis by Year
orders_by_year <- orders |>
  group_by(Year) %>%
  summarise(TotalOrders = n())

# Analysis by Month
orders_by_month <- orders |>
  group_by(Month) %>%
  summarise(TotalOrders = n())

# Analysis by Day
orders_by_day <- orders |>
  group_by(Day) |>
  summarise(TotalOrders = n())

# Visualize the results
knitr::kable(orders_by_year, caption = "orders_by_year")

# Display the results using DT::datatable
DT::datatable(
  orders_by_month,
  caption = "Orders by Month",
  options = list(
    lengthMenu = c(5, 10),
    pageLength = 5,
    searching = TRUE,
    ordering = TRUE
  ),
  style = "bootstrap",
  class = "table table-bordered table-hover"
)

DT::datatable(
  orders_by_day,
  caption = "Orders by Day",
  options = list(
    lengthMenu = c(5, 10),
    pageLength = 5,
    searching = TRUE,
    ordering = TRUE
  ),
  style = "bootstrap",
  class = "table table-bordered table-hover"
)
```

***

# Analyzing Monthly Trends: Rolling Average of Freight Costs
```{r, message=FALSE, warning=FALSE}

orders$OrderDate <- as.Date(orders$OrderDate)

# Function to calculate rolling average
calculate_rolling_average <- function(data, column, window_size) {
  data |>
    arrange(OrderDate) |>
    mutate(RollingAvg = zoo::rollapply(column, width = window_size, FUN = mean, align = "right", fill = NA))
}

# Calculate rolling average for Freight column with a window size of 3
orders <- calculate_rolling_average(orders, column = orders$Freight, window_size = 3)

# Plotting the trend using a rolling window on a monthly basis
plot_ly(orders, x = ~OrderDate, y = ~RollingAvg, type = 'scatter', mode = 'lines+markers', line = list(color = 'blue'), marker = list(color = 'red')) |>
  layout(title = "Trend of Rolling Average Freight Costs Over Time",
         xaxis = list(title = "Order Date"),
         yaxis = list(title = "Rolling Average Freight Cost"))

# Plotting time series plot
plot_ly(orders, x = ~OrderDate, y = ~Freight, type = 'scatter', mode = 'lines+markers', line = list(color = 'green'), marker = list(color = 'orange')) |>
  layout(title = "Time Series Plot of Freight Costs",
         xaxis = list(title = "Order Date"),
         yaxis = list(title = "Freight Cost"))

# Plotting boxplot to visualize monthly variation with plotly
plot_ly(data = orders, x = ~as.character(format(OrderDate, "%Y-%m")), y = ~Freight, type = 'box') |>
  layout(title = "Monthly Boxplot of Freight Costs",
         xaxis = list(title = "Month", tickangle = 45, categoryorder = "array", categoryarray = unique(as.character(format(orders$OrderDate, "%Y-%m")))),
         yaxis = list(title = "Freight Cost"))
```

***

# Customer Distribution by Region             
```{r, message=FALSE, warning=FALSE}
customers |>
  group_by(Region) |>
  summarise(NumCustomers = n()) |>
  plot_ly(labels = ~Region, values = ~NumCustomers, type = 'pie') |>
  layout(title = "Customer Distribution by Region")
```

***

# Analyzing Trends Over Time: Rolling Averages of Order Quantity by Category
```{r, message=FALSE, warning=FALSE}
# Use left_join for merging tables
merged_data <- left_join(orders, orderdetails, by = "OrderID") |>
  left_join(products, by = "ProductID")

# Function to calculate rolling average
calculate_rolling_average <- function(data, window_size) {
  data %>%
    arrange(OrderDate) |>
    group_by(CategoryID) |>
    mutate(RollingAverage = zoo::rollmeanr(Quantity, k = window_size, fill = NA)) |>
    ungroup()
}

# Set the rolling window size (e.g., 30 days for a month)
window_size <- 30

# Calculate rolling average
result <- calculate_rolling_average(merged_data, window_size)

# Plot the rolling average using Plotly with increased width
rolling_average_plot <- result |>
  plot_ly(x = ~OrderDate, y = ~RollingAverage, color = ~as.factor(CategoryID), type = 'scatter', mode = 'lines') |>
  layout(title = "Rolling Average Order Quantity Over Time by Category",
         xaxis = list(title = "Order Date"),
         yaxis = list(title = "Rolling Average Quantity"),
         showlegend = TRUE,
         width = 800)  # Adjust the width as needed

# Display the Plotly plot
rolling_average_plot
```

***

# Customer-wise Orders and Shipments
```{r, message=FALSE, warning=FALSE}
# Full join Customers with Orders
full_joined_customers_orders <- full_join(customers, orders, by = "CustomerID")

# Count the number of orders and shipments for each customer
customer_orders_shipments <- full_joined_customers_orders |>
  group_by(CustomerID, CompanyName) |>
  summarise(OrdersCount = n(), ShipmentsCount = sum(!is.na(ShippedDate)))

# Select the top 10 customers based on the number of orders
top_customers_orders <- customer_orders_shipments |>
  arrange(desc(OrdersCount)) |>
  head(10)

# Display the Supplier Contribution
knitr::kable(top_customers_orders, caption = "Customer Orders Shipments")
```

***

# Analyzing Monthly Trends: Interactive Rolling Average for Freight Costs
```{r, message=FALSE, warning=FALSE}
# Make sure 'OrderDate' is of type Date
orders$OrderDate <- as.Date(orders$OrderDate)

# Function to calculate rolling average
calculate_rolling_average <- function(data, column, window_size) {
  data |>
    arrange(OrderDate) |>
    mutate(RollingAvg = zoo::rollapply(column, width = window_size, FUN = mean, align = "right", fill = NA)) |>
    select(OrderDate, RollingAvg)
}

# Calculate rolling average for Freight column with a window size of 30
rolling_avg_table <- calculate_rolling_average(orders, column = orders$Freight, window_size = 3)

plot_ly() |>
  add_lines(data = rolling_avg_table, x = ~OrderDate, y = ~RollingAvg, type = 'scatter', mode = 'lines', name = 'Rolling Avg') |>
  add_lines(data = orders, x = ~OrderDate, y = ~Freight, type = 'scatter', mode = 'lines', name = 'Freight Cost') |>
  layout(title = "Interactive Time Series Plot with Rolling Average",
         xaxis = list(title = "Order Date"),
         yaxis = list(title = "Freight Cost"))
```

***
